---
title: "README"
output: html_document
date: "2024-10-25"
---

# ADAM: Adaptive Differential Abundance Method

'ADAM' is an R package designed to implement an Adaptive approach for Differential Abundance Analysis. It provides a streamlined workflow for analyzing data, selecting the best-performing method, and generating robust results.

## Installation

To install the `ADAM` package directly from GitHub, follow these steps:

### 1. Ensure you have the `devtools` package installed:

```{r}
#install.packages("devtools")
```

### 2. Install `ADAM` from GitHub and load ADAM package:

```{r}
devtools::install_github("lucp9827/ADAM")
library(ADAM)
```

### Dependencies

The `ADAM` package depends on several external packages. These are automatically installed when you install `ADAM`, but ensure you have the following available in your R environment: - `dplyr` - `signtrans` - `phyloseq` - `ALDEx2` - `ANCOMBC` - `DESeq2` - `MicrobiomeStat` - `RioNorm2` - `corncob` - `microbiome`

#### You can install any missing dependencies manually:

```{r}
#install.packages(c("dplyr", "signtrans", "phyloseq"))

```

#### For Bioconductor packages (e.g., `ALDEx2`):

```{r}
#if (!requireNamespace("BiocManager", quietly = TRUE)) {
#  install.packages("BiocManager") BiocManager::install(c("ALDEx2", "ANCOMBC", "DESeq2"))}
```

## Using the ADAM_main Function

The main function of the package, `ADAM_main`, performs the full ADAM workflow. Here's a step-by-step guide to use it.

We use example data for scenario 3.3, which has 100 simulated dataset using the SPSimseq framework. These datasets have 250 taxa, 75 samples per group with 10% DA taxa which has been selected using a threshold of LFC=1.5 and in the setting with low sparisty. You can replace this with your dataset.

```{r}
load("simsB_3.3_p.RData")
data <- sim.data.bulk.p_B[[1]]
```

Parameters: - `norm_method`: Normalization method (default: `"TSS"`). - `perc_strata`: Proportion of taxa of the original data in strata (default: `0.2`). - `OTU_replace`: Logical; Sampling with replacement (default: `TRUE`). - `threshold_value`: Logical; whether to use adaptive thresholding (default: `TRUE`). - `alpha`: Significance level for FDR control (default: `0.05`). - `B`: Number of bootstrap/permutation datasets (default: `2`). - `path`: File path to save results (default: current working directory). - `dataset`: Numeric identifier for the dataset (default: `1`).

Note: Create a Folder in your working directory titled 'Variations', which is used to store the variations of the Test datasets that are created. Intermediate results (Test dataset, performance of DA methods on Variations of Test datasets, raw p-values of each test) are saved in your working directory.

### Running the Function

```{r}
library(ADAM)

# Defining Parameters
norm_method <- "TSS" 
perc_strata <- 0.2 
OTU_replace <- TRUE 
threshold_value <- TRUE 
alpha <- 0.05 
B <- 1 
path <- getwd() 
dataset <- 1

final_results <- ADAM_main( data = data, norm_method = norm_method, perc_strata = perc_strata, OTU_replace = OTU_replace, threshold_value = threshold_value, alpha = alpha, B = B, path = path, dataset = dataset )
```

The `ADAM_main` function returns a list:

-   `final_res`: Performance metrics (Type I error rate, sensitivity, FDR).

-   `method`: The statistical method chosen as best DA method

-   `test`: Raw results of the method applies (statistic, raw pvalues, adjusted pvalues (BH-procedure))

### Troubleshooting ---------------------------------------------------------

-   **Missing Dependencies**: Ensure all required packages are installed (see the Dependencies section).

-   **Data Format**: Input data should conform to the expected format for phyloseq objects, with 'group' used in the sample data information to determine group allocation

#### Contributing

Contributions are welcome! Please submit issues or pull requests to improve the package or its documentation. This package is still being actively developed and optimalised.
